$p$-adic precision

### 2014-02-14 padicprec.tex
Working on introduction

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{Bostan} use Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{Gaudry} use $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{Kedlaya} uses $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Survent \cite{Lercier} compute isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

@@

In the $p$-adic world, precision tracking using intervals is much more feasible.
Even a long sequence of operations may not sacrifice any precision.  Intervals
allow number theorists to provably determine a result modulo a given power of $p$,
and the Gaussian distributions of measurement error over $\R$ have no direct analogue
over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.

Many usual softwares (\emph{e.g.} {\tt magma} and {\tt sage}) include 
primitives to handle $p$-adic numbers. Since $p$-adic numbers are by 
nature infinite, one needs to truncate them up to some finite level (we 
generally say \emph{finite precision}) in order to manipulate them on a 
computer. The usual technics to achieve this are inspired from case of 
real numbers: the computer works with ``truncated $p$-adic numbers'' 
represented as $x + O(p^n)$ and performs all basic operations (addition, 
product) on this representation. For instance, it computes: 
$$(x + O(p^n)) + (y + O(p^m)) = (x+y) + O(p^{\min(n,m)}).$$
However, in many important cases (\emph{e.g.} those discussed just 
above), this approach is too naive since it creates artificially 
important losses of precision (see \S\ref{ssec:stepbystep}). 
It is the so-called \emph{numerical instability} issue.

The aim of the current paper is to provide a systematical theorical 
study of $p$-adic precision (and more generally nonarchimedian 
precision) and to derive from this study new methods to handle 
precision, that turns out to give nearly optimal results in a quite 
large setting.

\section{Motivations and aims}

>>>>

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{Bostan} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{Gaudry} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

@@

In the $p$-adic world, precision tracking using intervals is much more feasible.
Even a long sequence of operations with such elements may not sacrifice any precision.  Intervals
allow number theorists to provably determine a result modulo a given power of $p$,
and the Gaussian distributions of measurement error over $\R$ have no direct analogue
over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
in implementations of $p$-adic numbers.  The mathematical software packages
Sage \cite{sage}, PARI \cite{pari} and Magma \cite{magma} all include $p$-adic elements
that track precision in this way.

The approach of propagating precision with each arithmetic operation works well, but does
sometimes underestimate the known precision of a result, as we will discuss in Section
\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
implementations of polynomials, vector spaces, matrices and power series.  The practice
of storing the precision within each entry is not flexible enough for all applications.
Sometimes only a rough accounting of precision is needed, in which case storing
and computing the precision of each entry in a large matrix needlessly consumes
space and time.  Conversely, recording the precision of each entry does not allow
a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
quadratic polynomial $f$.

For a vector space $V$ over $\Qp$, we propose that the fundamental object used to
store the precision of an element should be a $\Zp$-lattice $H \subset V$.  By using
general lattices one can eliminate needless loss of precision.  Moreover, specifying
the precision of each entry or recording a fixed precision for all entries can both
be interpreted in terms of lattices.  In Section \ref{sec:prec-proposal} we detail our
proposal for how to represent the precision of an element of a vector space.

\todofor{David}{Finish summarizing sections}

While we have used $\Qp$ throughout this paper for concreteness, all of our results
and suggestions apply to arbitrary complete discrete valuation rings.

\section{Tracking Precision} \label{sec:prec-proposal}
